% ---------------------------------------------
\chapter{Performing polygons\textasciitilde{}}
\label{sec: performance}
\markboth{}{Weird Polygons and Hand Noises}
\epigraph{\emph{quote}}{\citep[]{bilbow2022}}
% ---------------------------------------------
%              _                            /\/|
%  _ __   ___ | |_   _  __ _  ___  _ __  __|/\/
% | '_ \ / _ \| | | | |/ _` |/ _ \| '_ \/ __|
% | |_) | (_) | | |_| | (_| | (_) | | | \__ \
% | .__/ \___/|_|\__, |\__, |\___/|_| |_|___/
% |_|            |___/ |___/
% ---------------------------------------------
\section{Developing an AR Performance Practice} \label{sec: performance-developing}
As a direct consequence of witnessing participants enjoyment and play with polaris\textasciitilde{}, I recognised for the first time looking from the outside in, that the system had a larger propensity for fostering learning, virtuosity, and depth of expression than I had originally thought; if only the experience was slightly more complex, the interactions more \textit{nuanced}. The gestures, play, and expression by participants led to a kind of quasi-transhumanist dance, one that struck me as a technologically mediated dialogue between hybrid self and hybrid environment, a delicate balance of agency.

Until this point, I had been primarily focused on the medium as used for the composition of musical pieces e.g. \hyperref[sec: area]{area\textasciitilde{}}, or of installation-like experiences \hyperref[sec: polaris]{polaris\textasciitilde{}}. Performance with AR had remained abstract, especially as, formally, its not my preferred mode of musical expression. However, the balance of agency between participant and system, between self and environment, presented itself during the evaluation of polaris\textasciitilde{} to be an area ready for exploration through performance. 

Most of the software and hardware I had used in the development of polaris\textasciitilde{} was readily transferable to the context of performance. There was no need to change from the combination of PureData, LibPdIntegration, and Unity. However, performance did confront me with new considerations: 
\begin{itemize}
    \item \textit{What elements of my experience would need to be shared with an audience?} 
    \item \textit{What role would my gestures have in describing the experience that I was by now intimately aware of, but would be strange for an audience?} 
    \item \textit{Would I share the visual experience as well as the auditory experience?} 
\end{itemize}
The following chapter outlines my exploration of a performance called \textit{Weird Polygons and Hand Noises}, a fifteen-minute experimental audiovisual AR improvisation.



% ---------------------------------------------
\section{Scene Composition} \label{sec: performance-composition}
\textit{Weird Polygons and Hand Noises} is the name given to the set of AR instruments I first performed live. The ``Scene'' as it will be referred to in this chapter encompasses the 

Learning => Improvisation


% ---------------------------------------------
\section{Technical Setup} \label{sec: performance-setup}




% ---------------------------------------------
\section{Performances} \label{sec: performance-performances}
\subsection{The Rosehill} \label{sec: performance-performances-rosehill}
\subsection{The Attenborough Centre for Creative Arts} \label{sec: performance-performances-acca}



% ---------------------------------------------
\section{Demonstrations} \label{sec: performance-demonstrations}
\subsection{MAH Doctoral Conference} \label{sec: performance-demonstrations-mah}
\subsection{SHL Make \& Create} \label{sec: performance-demonstrations-shl}